The role of animal cognition in shaping resource distribution

### 2023-07-06 Text/Article_ModelSeedDispersal.Rmd
Update after finalisation - first sent to colleagues

#I follow: https://stackoverflow.com/questions/52918716/authors-and-affiliations-in-the-yaml-of-rmarkdown for author display
author: 
  - name: Benjamin Robira
    email: benjamin.robira@normalesup.org
    institute: FEM
    correspondence: true
institute:
  - FEM: Animal Ecology Unit, Research and Innovation Centre, Fondazione Edmund Mach, San Michele all'Adige, TN, Italy

@@

urlcolor: gray
filecolor: gray
linkcolor: gray
fontsize: 12pt

header-includes:
  #Emoticone
  - \usepackage{fontawesome5}
  #Customise page
  - \usepackage{fancyhdr}
  - \usepackage{lastpage} #Number of lastpage
  #- \pagenumbering{gobble} #for no page numbering
  - \usepackage{pdfpages}
  # #Underline text
  # - \usepackage{ulem}
  #For graybox
  - \usepackage{tcolorbox}
  #For pictures, drawing...
  - \usepackage{tikz}
  #For title sections
  - \usepackage{titlesec}
  #Lettrine
  - \usepackage{lettrine}
  #Customise legend figure
  - \usepackage[rightcaption]{sidecap}
  - \usepackage{graphicx}
  - \usepackage{subcaption}
  - \usepackage{floatrow}
  - \usepackage{caption}
  #Multiple columns
  - \usepackage{multicol}
  #To have lines around title
  - \usepackage{titling}
  - \pretitle{
    \begin{flushleft}
    \rule[-0.15in]{0.25\linewidth}{0.8ex}
    \vspace{-0.8ex}
    \hrule
    \vspace{0.3in}
    \begin{LARGE}
    \noindent
    \textbf
    }
  - \posttitle{
    \end{LARGE}\newline
    \rule[-0.15in]{0.25\linewidth}{0.8ex}
    \hrule
    \end{flushleft}
    \vspace{0.2in}}
  - \usepackage{lettrine}  
  #- \usepackage[nolists, nomarkers,tablesfirst]{endfloat} # For figures and tables at end
  - \usepackage{lineno} # For line numbering
  - \linenumbers # For line numbering
  #- \usepackage{setspace}\doublespacing
  - \usepackage{fontawesome} #for fa symbols
  - \usepackage{tcolorbox}
  #- \pagenumbering{gobble} #for no page numbering
  - \setlength{\parskip}{0em} #to remove line gaps
  #- \setlength{\parindent}{2em} #to indent new paragraph
  - \DeclareUnicodeCharacter{2212}{-}
  - \usepackage{caption}
  - \captionsetup[figure]{font=small}
  - \newcommand{\beginsupplement}{
    \setcounter{table}{0}  
    \renewcommand{\thetable}{A\arabic{table}} 
    \setcounter{figure}{0} 
    \renewcommand{\thefigure}{A\arabic{figure}}
    } 
---

@@

:::: {.graybox data-latex=""}

**Abstract |** In the wildlife theater, an evolutionary play takes place in an unchanging setting, or so current hypotheses on the evolution of spatiotemporal cognition of plant-eater animals assume. If these animals are likely to have their spatiotemporal cognition shaped by the distribution of resources, their movement may shape the way plant resources are distributed by dispersing their seeds too. This could create eco-evolutionary feedback loops between animal spatiotemporal cognition and the distribution patterns of plant resources. Using an agent-based model simulating the foraging behaviour of a seed disperser endowed with spatiotemporal knowledge of resource distribution, we investigated whether resource spatiotemporal patterns could be influenced by the level of cognition involved in foraging. This level of cognition represented how well resource location and phenology were predicted by the agent. We showed that seed dispersers could shape the long-term resource distribution with the formation of materialised resource routes. This stemmed from the conjunction of two forces: competition for space between trees and a seed-dispersal forager moving from tree to tree based on spatiotemporal memory. However, the engineering of the resource landscape did not substantially affect the selection gradient that, in turn, could apply to cognition. Altogether, the results emphasise that cognition is a cause and a consequence of resource heterogeneity.

@@

<!-- Using an agent-based model simulating the foraging behaviour of a seed disperser endowed with spatiotemporal knowledge of resource distribution, we investigated whether resource spatiotemporal patterns could be influenced by the level of cognition involved in foraging. This level of cognition represented how well resource location and phenology were predicted by the agent. We showed that seed dispersers could shape the long-term resource distribution with the formation of materialised resource routes. This stemmed from the conjunction of two forces: competition for space between trees and a seed-dispersal forager moving from tree to tree based on spatiotemporal memory. However, the engineering of the resource landscape did not affect the selection gradient that, in turn, could apply to cognition.  -->

@@

\initial{\textcolor{black}{P}}erhaps because animal mobility contrasts with plant sessility, hypotheses on the evolution of plant foragers' spatiotemporal cognition generally assume that moving foragers evolve in an unchanging environment over the long term [@rosati2017foraging]. However, while foragers' cognition is likely to be shaped by the distribution of resources [@milton1981distribution; @grove2013evolution; @robiratemporalmemory; @bracis2015memory; @boyer2010modelling], the characteristics of the resources themselves could be the result of forager-induced selection [@valenta2020dispersal; @lomascolo2010signal]. In the tropical rainforests, for example, the cognition of frugivorous primates has probably been shaped by the challenge of the ephemerality and dispersal of their food [@zuberbuhler2010foraging; @trapanese2019review], supporting complex [@teichroeb2019navigation; @trapanese2019whatwhere; @janson2016capuchins; @janson1992measuring; @janmaat2013mistake; @robirarandomwalk] and flexible [@trapanese2022flexibility; @janmaat2006primates] movement heuristics. In turn, primates remember and forage for the most rewarding trees [@ban2016fattyfruit; @ban2014highreward; @florchinger2010fruit]. The latter are thus more likely to have their seeds dispersed, driving plant and landscape evolution [@chapman2013primates].

|   The way seeds are dispersed contributes to the emerging pattern of resource distribution [@vittoz2007seed]. Most of the world's plants [@tiffney1995angiosperm] and especially rainforest trees [main and supplementary material: @gottsberger1983dispersal; @bagchi2011spatial] are dispersed by animals (zoochory). Compared to other modes of dispersal, such as wind (anemochory) or gravity (barychory), zoochory induces a more patchy distribution of adult plants [@seidler2006seed]. This is because animals may tend to revisit memorised sites regularly, and to some extent exclusively [@john2016memoryseed; @addis2017causes]. Furthermore, the imperfection and limited size of spatial memory [@fagan2013spatial] is expected to lead to different movement patterns, and therefore potentially different patterns of resource distribution [@cortes2013integrating]. By being affected by cognition and by affecting resource dispersal, animal movement could thus lead to eco-evolutionary feedback loops between animal cognition and resource dispersal [@john2016cognitionmeetdispersal]. In other words, cognition could be both a cause and a consequence of resource heterogeneity.

|   To illustrate the possible retroactions between cognition and resource distribution, let's consider an extreme speculative scenario in which seed-dispersing frugivores are freely foraging (Figure \@ref(fig:predictions)). At start, assuming a parcimonious homogeneous distribution, long-term spatiotemporal memory is only marginally advantageous [@grove2013evolution]. As it is nonetheless, cognition should be selected for. Through zoochory, deliberate targeting of some resource patches over others may induce a change in resource distribution over time (e.g. increasing patchiness), eventually making cognition more and more advantageous. Gradually, by repeatedly targeting the same trees, the routes used by the foragers may begin to materialise through the seeds scattered along the way. However, in doing so, foragers who can only passively follow these routes and have survived may also begin to become very efficient: they should not even have to bear the cost of high cognitive ability, putting them at an advantage over foragers who bear the burden of cognition [@raichle2006brain]. While the pattern of resource distribution can be maintained because the routes are used equally by all foragers [but through two different mechanisms: long-term memory vs. taxis, @benhamou1995comparative; @benhamou1992distinguishing], cognition could even start to be counter-selected.

|   While some facets of this scenario may be rooted in empirical evidence [e.g. with forest elephants, @campos2011megagardeners; @blake2004fruit; @blake2009forest], most of this scenario remains speculative. The aim of this study is thus to put such a scenario to the test. Using an agent-centred model simulating the movement of seed-dispersing frugivores with different cognitive abilities, we sought to investigate (a) whether the spatial arrangement of resource varies with the cognitive level and movement behaviour of the simulated agents, and (b) whether this in turn affects the selection pressure on the agents' cognition.

@@

The main objective of the model was to simulate the foraging behaviour of a single agent in an environment dynamical both on the short term, through an interplay between resource growth and depletion, and on the long term, due to seed dispersal. For simplicity, we will refer to the agent as the forager, and to resource patches as fruit trees, but this model should find larger resonance. The model was implemented in the *Rcpp* language in *R* software [`r extractRversion()`, @Rsoftware]  based on the *Rcpp* package [@Rcpp1; @Rcpp2; @Rcpp3].

@@

We modelled the environment as a flat square map of side `r mapSize` arbitrary spatial units (su). Within this map, `r numberTrees` trees were initially homogeneously distributed (uniform distribution of their $x$ and $y$ coordinates, resulting in a Poisson distribution of tree density).

@@

Trees could reproduce and disperse via fruit eaten by the agent. Once the agent ate on ripe fruit, seeds could be dispersed once up to `r timeDelayForDispersal` tu after ingestion, with a probability of `r DispersalProbability/timeDelayForDispersal` tu$^{-1}$, independent of the quantity of fruit ingested, for simplicity. Thus, all fruit eaten did not mandatorily led to dispersal. Dispersal was eventually successful only if some space was left available for the new tree. This means that no tree had to be at a distance lower than `r round(0.45 * sqrt(mapSize * mapSize / (numberTrees * pi)), digits = 2)` su. This number was obtained such as the exclusive "lifespace" by trees (a circular area of radius the aforementioned distance) was equal to 45% of the map area (but other percentages were tested, see [Route formation as a product of two forces][Route formation as a product of two forces: tree competition for space and agent movement step rule]). If it happened that there was available space for the seedling to grow, then the seedling location was sampled in the available space along the linear path linking the previous target. Otherwise, the seedling could not grow and died.

@@

#### Moving rules

The forager was initially randomly located within the map. It is fully aware of trees present within sensory range fixed to `r round(perceptualRange, digit=2)` su. This actually corresponds to the average nearest-neighbour distance which equates $\frac{1}{2\sqrt{\rho}}$, with $\rho$ the density of trees (here `r numberTrees/mapSize/mapSize`) at start of the simulation where trees are homogeneously distributed. It is also aware of trees part of its spatiotemporal long-term memory. To decide where to head next, the forager could either rely on its private knowledge (sensory or memory) or not. If all trees within its private knowledge did not yield any fruit (i.e., all having less than 0.001 fu), the forager was moving randomly. Random movements were performed by randomly choosing a direction (uniform distribution of the heading angle), while the step length was sampled within an exponential distribution of mean $\lambda_{step\,length}$ = `r exponentialRate`. An exponential distribution of step lengths is indeed classically observed in nature [REF]. Otherwise, the agent targeted the `r whatRule` tree which maximised the ratio between the predicted food quantity at the tree (which accounts for depletion history), and the distance to the tree. Nonetheless, the forager avoided to target trees recently visited within the previous `r noReturnTime` tu to avoid targeting recently depleted trees [@robiratemporalmemory].

|   Once the target had been chosen (or random point, if none), the forager moved straight to the first tree that should have been theoretically perceived *en route* to the initial destination point if any (provided they were not recently visited), or the destination otherwise. When this destination is reached, it then re-estimated the best tree to target and so on. Such a movement roughly mimics a beacon/landmark-based navigation, where individuals individuals move from beacon to beacon to reach a target [REF] and frames coherently with the tendency of individuals to monitor resource state while moving [REF]. After each moving bout time was updated based on the distance travelled, knowing that the forager moved at a speed of `r speed` su/tu. Then, the resource quantity available at each tree was updated.

@@

```{r modelParametersTable, include=TRUE}
library(kableExtra)
options(knitr.kable.NA = "-")
parametersValue[parametersValue[, 5] == "", 5] <- NA
knitr::kable(parametersValue, escape = TRUE, booktabs = TRUE, longtable = TRUE,
             caption = "Model parameters", align = "lllcccc") %>%
  kableExtra::kable_styling(latex_options = "striped") %>%
  kableExtra::landscape() %>%
  kable_styling(font_size = 7, full_width = FALSE)

@@

To assess whether different levels of cognition resulted in different organisation of the spatial distribution of trees at the end of the simulation we focused on three metrics describing the tree spatiotemporal patterns: the *patchiness*, the *alignment* and the *spatial autocorrelation of fruiting timing*.

@@

|   As the forager is set up to forage in a square map with reflective boundaries, seed dispersal may lead to the shrinkage of resource range towards the centre of the map (where recursions are inevitably the higher), an artefact not of interest there. This would contribute to inflate measures of patchiness. We therefore calculated a measure of shrinkage *s* as one minus the area covered by the 95% of the kernel distribution based on all tree locations divided by the map size. To account for variations in shrinkage in the different simulated scenarii [(see Appendix Material)][Variations in resource range shrinkage intensity], we corrected patchiness such as $P_c = P(1-s)$.

@@

<!-- Pseudo fractal dimension (i.e. the tendency of trees to organise in patterns insensitive to observation scale) was calculated based on a Geographically Weighted Fractal Analysis [Applying two fractal methods to characterise the local and global deviations from scale invariance of built patterns throughout mainland France] specifically developed for point patterns. In brief, for points, the fractal dimension corresponds to the slope of the regression of the logarithm of the number of points around a point of interest, as a function of the radius length to consider neighbouring points. To do so, we used the "gwfa" function of the *gwfa* package [@gwfa]. We fixed the circular neighbourhood (a measure of whether the scale is local, low radius compared to the map size, or global, large radius compared to the map size) to 200 su. The pseudo-fractal dimension corresponds to the average pseudo fractal dimension of randomly sampled points (limited to 500 within cells of side 200 su), for which we varied the counting scale (i.e. circular radius around each point of interest) from 20 su to 60 su (ten regularly spaced values in total). We use the "pseudo" prefix because we did not verify whether the calculated fractal dimension is statistically significant (i.e. if the regression slope, aimed to determine the fractal dimension, indeed offers a different fit from a constant). -->

@@

<!-- Reticulation (i.e. the tendency of trees to form reticulated networks) of the tree distribution was measured following Addis (2017) [https://repositories.lib.utexas.edu/bitstream/handle/2152/68127/ADDIS-DISSERTATION-2017.pdf?sequence=1]. This method is based on the Relative Neighbourhood Graph [RNG, Toussaint, G. T. 1980. The relative neighbourhood graph of a finite planar set. Pattern Recognition 12:261–268], which should reflect the intuitive perception of reticulation a human observer might have. In this graph, two points are connected by an edge if there is no third point that is closer to the two points than the two are between each other. Such a property thus drives the visualisation of "void areas" within the distribution, since it favours immediate close-distance edge. It thus likely reflects the "routes" that are paved by trees that a forager might tend to follow. -->
<!-- We constructed the RNG as an undirected and weighted graph, where the edge weight corresponded to the Manhattan distance between the two connected points ("rng" function of the *cccd* package [@cccd]). We followed Addis (2017) and defined the reticulation as   -->
<!-- \hfill -->

@@

<!-- where *D* is the diameter of the RNG. This diameter corresponds to the length of the longest of the shortest paths that connect two points within the RNG. Shortest paths between each pair of points of the RNG were obtained with the "shortest.path" function of the *igraph* package [@igraph]. *M* corresponds to the Manhattan distance between the pair of points whose shortest path equates to that of the diameter (or its average, if multiple pairs could lead to the maximum diameter length), while $M_{i,j}$ corresponds to the Euclidean distance between two points *i* and *j*. Despite we are not in a grid-based system (Addis, 2017), we used the Manhattan distance because we empirically observed that this enabled us distinguishing correctly between different spatial patterns (see Appendix [What distance to calculate reticulation?]) -->

@@

Routes are (curvi-)linear features. To assess the tendency of trees to form linear motifs, we quantified their alignment. For a given tree, alignment *a* was defined as one minus the minimum of the sine of the angles of the triangle formed by the tree of interest and the two closest trees. This varies from perfect alignment (value of 1), to no alignment (i.e., the minimum angle would be at worse 60° hence an alignment of ca. 0.134). Therefore, we used a corrected alignment $a_c$ varying from 0 (no alignment) to 1 (perfect alignment) such as

@@

We defined the general alignement for a given environmental map as the opposite of the skewness of alignment values of all trees in the map. Positive values indicate the existence of linear motifs while negative values indicate no obvious linear motifs. Random distributions (homogeneous or heterogeneous) should show moderate positive values, while a route pattern should have a high positive value [(see Appendix Material)][Using patchiness and alignment to characterise spatial point patterns].

@@

Spatial autocorrelation of fruiting timing (i.e. the tendency of fruiting date of trees to more closely match that of neighbours than of distant trees) was calculated as the Geary Index, \textit{C} [@geary1954contiguity] applied to the start of fruiting dates. Since the date of start of fruiting corresponds to a circular variable, we followed [@schmal2017moran] to adapt existing functions of the *ape* package [@ape] to circular variables. Geary's \textit{C} was therefore defined as

@@

where $N$ is the number of trees, $t$ is the fruiting date (in radian), $d(t_1,t_2) = atan2(sin(t_1 - t_2), cos(t_1 - t_2))$, $\bar{t} = atan2(1/N\sum_isin(t_i), 1/N\sum_icos(t_i))$. We used the inverse of the Euclidean distance between tree locations as weights. Values of $C$ close to 1 indicate absence of spatial autocorrelation. Values close to 0 indicate negative spatial autocorrelation. Values larger than 1 indicate positive spatial autocorrelation.

@@

For each simulation, we verified whether a routine behaviour emerged and paralleled the variations in tree distribution. We characterised routine holding on the notion of conditional entropy following [@riotte2016randomness] using the series of the id of trees targeted during five supplementary seasons in which dispersal could not occur. We computed the routine possibly up to only a first order conditional entropy, as computing routine indices is computationally demanding and a first order is expecting to be the most informative to describe a perfect traplining behaviour [@riotte2016randomness]. The routine index varies from 0 (no routine) to 1 (perfect traplining).

>>>>

#I follow: https://stackoverflow.com/questions/52918716/authors-and-affiliations-in-the-yaml-of-rmarkdown for author display
author: 
  - name: Benjamin Robira \orcidlink{0000-0002-3168-6573}
    email: benjamin.robira@normalesup.org
    institute: FEM
    correspondence: true
institute:
  - FEM: Animal Ecology Unit, Research and Innovation Centre, Fondazione Edmund Mach, San Michele all'Adige, TN, Italy

@@

urlcolor: gray
filecolor: gray
link-citations: yes
linkcolor: gray
citecolor: gray
fontsize: 11pt

header-includes:
  #Lato font
  - \usepackage[default]{lato}
  - \usepackage{lipsum}% for dummy text
  - \renewcommand{\mddefault}{l}
  #Orcid link
  - \usepackage{orcidlink}
  #Emoticone
  - \usepackage{fontawesome5}
  #Customise page
  - \usepackage{fancyhdr}
  - \usepackage{lastpage} #Number of lastpage
  #- \pagenumbering{gobble} #for no page numbering
  - \usepackage{pdfpages}
  # #Underline text
  # - \usepackage{ulem}
  #For graybox
  - \usepackage{tcolorbox}
  #For pictures, drawing...
  - \usepackage{tikz}
  #For title sections
  - \usepackage{titlesec}
  #Lettrine
  - \usepackage{lettrine}
  #Customise legend figure
  - \usepackage[rightcaption]{sidecap}
  - \usepackage{graphicx}
  - \usepackage{subcaption}
  - \usepackage{floatrow}
  - \usepackage{caption}
  #Multiple columns
  - \usepackage{multicol}
  #To have lines around title
  - \usepackage{titling}
  - \pretitle{
    \begin{flushleft}
    \rule[-0.15in]{0.25\linewidth}{0.8ex}
    \vspace{-0.8ex}
    \hrule
    \vspace{0.3in}
    \begin{LARGE}
    \noindent
    \textbf
    }
  - \posttitle{
    \end{LARGE}\newline
    \rule[-0.15in]{0.25\linewidth}{0.8ex}
    \hrule
    \end{flushleft}
    \vspace{0.2in}}
  - \usepackage{lettrine}  
  #- \usepackage[nolists, nomarkers,tablesfirst]{endfloat} # For figures and tables at end
  - \usepackage{lineno} # For line numbering
  - \linenumbers # For line numbering
  #- \usepackage{setspace}\doublespacing
  - \usepackage{fontawesome} #for fa symbols
  - \usepackage{tcolorbox}
  #- \pagenumbering{gobble} #for no page numbering
  - \setlength{\parskip}{0em} #to remove line gaps
  #- \setlength{\parindent}{2em} #to indent new paragraph
  - \DeclareUnicodeCharacter{2212}{-}
  - \usepackage{caption}
  - \captionsetup[figure]{font=small}
  - \newcommand{\beginsupplement}{
    \setcounter{table}{0}  
    \renewcommand{\thetable}{A\arabic{table}} 
    \setcounter{figure}{0} 
    \renewcommand{\thefigure}{A\arabic{figure}}
    } 
---

@@

:::: {.contouronlybox data-latex=""}

**Abstract |** In the wild, an evolutionary play takes place in an unchanging setting, or so current hypotheses on the evolution of spatiotemporal cognition of animals implicitly assume. If plant-eating vagile animals are likely to have their spatiotemporal cognition shaped by the distribution of plant resources, their movement may shape the way resources are distributed as well, by dispersing seeds. This could create eco-evolutionary feedback loops between animal spatiotemporal cognition and the distribution patterns of plant resources. Using an agent-based model simulating the foraging behaviour of a seed disperser endowed with spatiotemporal knowledge of resource distribution, I investigated whether resource spatiotemporal patterns could be influenced by the level of cognition involved in foraging. This level of cognition represented how well resource location and phenology were predicted by the agent. I showed that seed dispersers could shape the long-term resource distribution with the formation of materialised resource routes. This stemmed from the conjunction of two forces: competition for space between trees and a seed-dispersing forager moving from tree to tree based on spatiotemporal memory. In turn, this engineering affected the benefits of spatiotemporal memory, hence the selection gradient that applies to cognition. Altogether, the results emphasise that foraging cognition is a cause and a consequence of resource heterogeneity.

@@

<!-- Using an agent-based model simulating the foraging behaviour of a seed disperser endowed with spatiotemporal knowledge of resource distribution, I investigated whether resource spatiotemporal patterns could be influenced by the level of cognition involved in foraging. This level of cognition represented how well resource location and phenology were predicted by the agent. I showed that seed dispersers could shape the long-term resource distribution with the formation of materialised resource routes. This stemmed from the conjunction of two forces: competition for space between trees and a seed-dispersal forager moving from tree to tree based on spatiotemporal memory. However, the engineering of the resource landscape did not affect the selection gradient that, in turn, could apply to cognition.  -->

@@

\initial{\textcolor{black}{H}}ypotheses about the evolution of vagile plant-eating foragers' spatiotemporal cognition (i.e. the ability to process, integrate and utilise information on resource distribution and phenology) generally assume that foragers evolve in an unchanging environment [@rosati2017foraging]. However, while foragers' cognition is likely to be shaped by the distribution of resources [@milton1981distribution; @grove2013evolution; @robiratemporalmemory; @bracis2015memory; @boyer2010modelling], the characteristics of the resources themselves could be the result of forager-induced selection [@lomascolo2010signal; @rojas2021combination]. In the tropical rainforests, for example, the cognition of frugivorous primates has probably been shaped by the challenge of the ephemerality and dispersal of their food [@milton1981distribution; @zuberbuhler2010foraging; @trapanese2019review], supporting complex [@teichroeb2019navigation; @trapanese2019whatwhere; @janson2016capuchins; @janson1998experimental; @janmaat2013mistake; @robirarandomwalk] and flexible [@trapanese2022flexibility; @janmaat2006primates] movement heuristics. In turn, primates remember and forage for the most rewarding trees [@ban2016fattyfruit; @ban2014highreward; @florchinger2010fruit]. The latter are thus more likely to have their seeds dispersed, driving plant and landscape evolution [@chapman2013primates; @lambert1998evolutionary].

|   The way seeds are dispersed contributes to the emerging pattern of resource distribution [@vittoz2007seed]. Most of the world's plants [@tiffney1995angiosperm] and especially rainforest trees [@gottsberger1983dispersal; @bagchi2011spatial] are dispersed by animals (zoochory). Compared to other modes of dispersal, such as wind (anemochory) or gravity (gravichory), zoochory induces a more patchy distribution of adult plants [@seidler2006seed]. This is because animals may tend to revisit memorised sites regularly, if not exclusively [@john2016memoryseed; @addis2017causes]. Furthermore, the imperfection and limited size of spatial memory [@fagan2013spatial] is expected to lead to different movement patterns [@avgar2013empirically], and therefore potentially to different patterns of resource distribution [@cortes2013integrating]. By being affected by cognition and by affecting resource dispersal, animal movement could thus lead to eco-evolutionary feedback loops between animal cognition and resource dispersal [@borah2022studying]. In other words, foraging cognition could be both a cause and a consequence of resource heterogeneity.

|   To illustrate the possible retroactions between cognition and resource distribution, let's consider an extreme speculative scenario in which seed-dispersing frugivores are freely foraging (Figure \@ref(fig:predictions)). At start, assuming a parcimonious homogeneous distribution, long-term spatiotemporal memory is only marginally advantageous [@grove2013evolution]. As it is nonetheless, cognition should be selected for. Through zoochory, deliberate targeting of some resource patches over others may induce a change in resource distribution over time (e.g. increasing patchiness), eventually making cognition more and more advantageous. Gradually, by repeatedly targeting the same trees, the routes used by the foragers may begin to materialise through the seeds scattered along the way. However, in doing so, foragers who can only passively follow these routes and have survived may also begin to become very efficient: they should not even have to bear the cost of high cognitive ability, putting them at an advantage over foragers who bear the burden of cognition [@raichle2006brain; @burns2011costs]. While the pattern of resource distribution can be maintained because the routes are used equally by all foragers [but through two different mechanisms: long-term memory *vs* taxis, @benhamou1995comparative; @benhamou1992distinguishing], cognition could even start to be counter-selected.

|   While some facets of this scenario may be rooted in empirical evidence [e.g. with forest elephants, @campos2011megagardeners; @blake2004fruit; @blake2009forest], most of it remains speculative. The aim of this study is thus to put such a scenario to the test. Using an agent-based model simulating the movement of seed-dispersing frugivores with different cognitive abilities, I sought to investigate (a) whether and how the spatial arrangement of resource varies with the cognitive level and movement behaviour of the simulated agents, and (b) whether and how this in turn affects the selection pressure on the agents' cognition.

@@

The main objective of the model was to simulate the foraging behaviour of a single agent in an environment dynamical both on the short term, through an interplay between resource growth and depletion, and on the long term, due to seed dispersal. For simplicity, I will refer to the agent as the forager, and to resource patches as fruit trees, but this model should find larger resonance. The model was implemented in the *Rcpp* language in *R* software [`r extractRversion()`, @Rsoftware]  based on the *Rcpp* package [@Rcpp1; @Rcpp2; @Rcpp3].

@@

I modelled the environment as a flat square map of side `r mapSize` arbitrary spatial units (su). Within this map, `r numberTrees` trees were initially homogeneously distributed (uniform distribution of their $x$ and $y$ coordinates, resulting in a Poisson distribution of tree density).

@@

Trees could reproduce and disperse via fruit eaten by the agent. Once the agent ate on ripe fruit, seeds could be dispersed once up to `r timeDelayForDispersal` tu after ingestion, with a probability of `r DispersalProbability/timeDelayForDispersal` tu$^{-1}$, independent of the quantity of fruit ingested, for simplicity. Thus, all fruit eaten did not mandatorily led to dispersal. Dispersal was eventually successful only if some space was left available for the new tree: no tree had to be at a distance lower than `r round(0.45 * sqrt(mapSize * mapSize / (numberTrees * pi)), digits = 2)` su. This number was obtained such as the cumulative (when non overlapping) exclusive "lifespace" by trees (a circular area of radius the aforementioned distance) reached at best 45% of the map area (but other percentages were tested, see [Route formation as a product of two forces][Q3. Route formation as a product of two forces? Tree competition for space and forager movement step rule]). If it happened that there was available space for the seedling to grow, then the seedling location was sampled in the available space along the linear path linking the previous and current locations. Otherwise, the seedling could not grow and died.

@@

#### Movement rules

The forager was initially randomly located within the map. It is fully aware of trees present within sensory range fixed to `r round(perceptualRange, digit=2)` su. This actually corresponds to the average nearest-neighbour distance which equates $\frac{1}{2\sqrt{\rho}}$, with $\rho$ the density of trees (here `r numberTrees/mapSize/mapSize`) at start of the simulation where trees are homogeneously distributed. It is also aware of trees part of its spatiotemporal long-term memory. To decide where to head next, the forager could either rely on its private knowledge (sensory or memory) or not. If all trees within its private knowledge did not yield any fruit (i.e., all having less than 0.001 fu), the forager was moving randomly. Random movements were performed by randomly choosing a direction (uniform distribution of the heading angle), while the step length was sampled within an exponential distribution of mean $\lambda_{step\,length}$ = `r exponentialRate`. An exponential distribution of step lengths is indeed classically observed in nature [@benhamou2014scales]. Otherwise, the agent targeted the `r whatRule` tree which maximised the ratio between the predicted food quantity at the tree (which accounts for depletion history), and the distance to the tree. Nonetheless, the forager avoided to target trees recently visited within the previous `r noReturnTime` tu to avoid targeting recently depleted trees [@robiratemporalmemory].

|   Once the target had been chosen (or random point, if none), the forager moved straight to the first tree that should have been theoretically perceived *en route* to the initial destination point if any (provided they were not recently visited), or the destination otherwise. When this destination is reached, it then re-estimated the best tree to target and so on. After each moving bout time was updated based on the distance travelled, knowing that the forager moved at a speed of `r speed` su/tu. Then, the resource quantity available at each tree was updated.

@@

```{r modelParametersTable, include=TRUE}
source("../Scripts/R/0_Parameters.R")
library(kableExtra)
options(knitr.kable.NA = "-")
parametersValue[parametersValue[, 5] == "", 5] <- NA
knitr::kable(parametersValue, escape = TRUE, booktabs = TRUE, longtable = TRUE,
             caption = "Model parameters", align = "lllcccc") %>%
  kableExtra::kable_styling(latex_options = "striped") %>%
  kableExtra::landscape() %>%
  kable_styling(font_size = 7, full_width = FALSE)

@@

To assess whether different levels of cognition resulted in different organisations of the spatial distribution of trees at the end of the simulation I focused on three metrics describing the tree spatiotemporal patterns: the *patchiness*, the *alignment* and the *spatial autocorrelation of fruiting timing*.

@@

|   As the forager is set up to forage in a square map with reflective boundaries, seed dispersal may lead to the shrinkage of resource range towards the centre of the map (where recursions are inevitably the higher), an artefact not of interest there. This would contribute to inflate measures of patchiness. I therefore calculated a measure of shrinkage *s* as one minus the area covered by the 95% of the kernel distribution based on all tree locations divided by the map size. To account for variations in shrinkage in the different simulated scenarios [(see Appendix Material)][Variations in resource range shrinkage intensity], I corrected patchiness such as $P_c = P(1-s)$.

@@

<!-- Pseudo fractal dimension (i.e. the tendency of trees to organise in patterns insensitive to observation scale) was calculated based on a Geographically Weighted Fractal Analysis [Applying two fractal methods to characterise the local and global deviations from scale invariance of built patterns throughout mainland France] specifically developed for point patterns. In brief, for points, the fractal dimension corresponds to the slope of the regression of the logarithm of the number of points around a point of interest, as a function of the radius length to consider neighbouring points. To do so, I used the "gwfa" function of the *gwfa* package [@gwfa]. I fixed the circular neighbourhood (a measure of whether the scale is local, low radius compared to the map size, or global, large radius compared to the map size) to 200 su. The pseudo-fractal dimension corresponds to the average pseudo fractal dimension of randomly sampled points (limited to 500 within cells of side 200 su), for which I varied the counting scale (i.e. circular radius around each point of interest) from 20 su to 60 su (ten regularly spaced values in total). I use the "pseudo" prefix because I did not verify whether the calculated fractal dimension is statistically significant (i.e. if the regression slope, aimed to determine the fractal dimension, indeed offers a different fit from a constant). -->

@@

<!-- Reticulation (i.e. the tendency of trees to form reticulated networks) of the tree distribution was measured following Addis (2017) [https://repositories.lib.utexas.edu/bitstream/handle/2152/68127/ADDIS-DISSERTATION-2017.pdf?sequence=1]. This method is based on the Relative Neighbourhood Graph [RNG, Toussaint, G. T. 1980. The relative neighbourhood graph of a finite planar set. Pattern Recognition 12:261–268], which should reflect the intuitive perception of reticulation a human observer might have. In this graph, two points are connected by an edge if there is no third point that is closer to the two points than the two are between each other. Such a property thus drives the visualisation of "void areas" within the distribution, since it favours immediate close-distance edge. It thus likely reflects the "routes" that are paved by trees that a forager might tend to follow. -->
<!-- I constructed the RNG as an undirected and weighted graph, where the edge weight corresponded to the Manhattan distance between the two connected points ("rng" function of the *cccd* package [@cccd]). I followed Addis (2017) and defined the reticulation as   -->
<!-- \hfill -->

@@

<!-- where *D* is the diameter of the RNG. This diameter corresponds to the length of the longest of the shortest paths that connect two points within the RNG. Shortest paths between each pair of points of the RNG were obtained with the "shortest.path" function of the *igraph* package [@igraph]. *M* corresponds to the Manhattan distance between the pair of points whose shortest path equates to that of the diameter (or its average, if multiple pairs could lead to the maximum diameter length), while $M_{i,j}$ corresponds to the Euclidean distance between two points *i* and *j*. Despite I are not in a grid-based system (Addis, 2017), I used the Manhattan distance because I empirically observed that this enabled us distinguishing correctly between different spatial patterns (see Appendix [What distance to calculate reticulation?]) -->

@@

Routes are (curvi-)linear features. To assess the tendency of trees to form linear motifs, I quantified their alignment. For a given tree, alignment *a* was defined as one minus the minimum of the sine of the angles of the triangle formed by the tree of interest and the two closest trees. This varies from perfect alignment (value of 1), to no alignment (i.e., the minimum angle would be at worse 60° hence an alignment of ca. 0.134). Therefore, I used a corrected alignment $a_c$ varying from 0 (no alignment) to 1 (perfect alignment) such as

@@

I defined the general alignement for a given environmental map as the opposite of the skewness of alignment values of all trees in the map. Positive values indicate the existence of linear motifs while negative values indicate no obvious linear motifs. Random distributions (homogeneous or heterogeneous) should show moderate positive values, while a route pattern should have a high positive value [(see Appendix Material)][Using patchiness and alignment to characterise spatial point patterns].

@@

Spatial autocorrelation of fruiting timing (i.e. the tendency of fruiting date of trees to more closely match that of neighbours than of distant trees) was calculated as the Geary Index, \textit{C} [@geary1954contiguity] applied to the start of fruiting dates. Since the date of start of fruiting corresponds to a circular variable, I followed [@schmal2017moran] to adapt existing functions of the *ape* package [@ape] to circular variables. Geary's \textit{C} was therefore defined as

@@

where $N$ is the number of trees, $t$ is the fruiting date (in radian), $d(t_1,t_2) = atan2(sin(t_1 - t_2), cos(t_1 - t_2))$, $\bar{t} = atan2(1/N\sum_isin(t_i), 1/N\sum_icos(t_i))$. I used the inverse of the Euclidean distance between tree locations as weights. Values of $C$ close to 1 indicate absence of spatial autocorrelation. Values close to 0 indicate negative spatial autocorrelation. Values larger than 1 indicate positive spatial autocorrelation.

@@

For each simulation, I verified whether a routine behaviour emerged and paralleled the variations in tree distribution. I characterised routine holding on the notion of conditional entropy following [@riotte2016randomness] using the series of the id of trees targeted during five supplementary seasons in which dispersal could not occur. I computed the routine possibly up to only a first order conditional entropy, as computing routine indices is computationally demanding and a first order is expecting to be the most informative to describe a perfect traplining behaviour [@riotte2016randomness]. The routine index varies from 0 (no routine) to 1 (perfect traplining).

### Research questions and associated scenarios
